---
title: "Example analysis"
author: "Kening Jiang"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("tidyverse")
library("tidytuesdayR")
library("here")
library("table1")
library("ggpubr")
library("purrr")
```

# Research question

Examine hotel cancellations and factors associated with hotel cancellations in the United States between 2015 and 2017.

# Origin data

The dataset is included in [TidyTuesday, 2020-02-11](https://github.com/rfordatascience/tidytuesday/tree/master/data/2020/2020-02-11#readme). The original dataset comes from an open hotel booking demand dataset from [Antonio, Almeida and Nunes, 2019](https://www.sciencedirect.com/science/article/pii/S2352340918315191#f0010). This dataset includes bookings of resort hotels or city hotels between the 1st of July of 2015 and the 31st of August 2017, including bookings that effectively arrived and bookings that were canceled.

# Data dictionary

[Variables in Hotels](https://github.com/rfordatascience/tidytuesday/tree/master/data/2020/2020-02-11#data-dictionary)

# Data management

## Set up the dataset
```{r data, warning=FALSE, message=FALSE, results="asis"}
# tests if a directory named "data" exists locally
if (!dir.exists(here("data"))) {
    dir.create(here("data"))
}

if (!file.exists(here("data", "hotels.RDS"))) {
    tuesdata <- tidytuesdayR::tt_load("2020-02-11")
    hotels <- tuesdata$hotels
    # save the files to RDS object
    saveRDS(tuesdata$hotels, file = here("data", "hotels.RDS"))
}

# Load the dataset
hotels<- readRDS(here("data", "hotels.RDS"))
```

## Data cleaning
```{r dataclean, warning=FALSE, message=FALSE}
# Filter for USA hotel bookings and select relevant variables
hotels_clean<- 
  hotels %>%
  filter(country=="USA") %>%
  select(-c('arrival_date_week_number','arrival_date_day_of_month', 'meal','country', 'market_segment', 'distribution_channel', 
            'reserved_room_type', 'assigned_room_type', 'agent', 'company', 'days_in_waiting_list', 'required_car_parking_spaces', 
            'total_of_special_requests', 'reservation_status', 'reservation_status_date'))

# Drop the observations with missing values
hotels_clean <- na.omit(hotels_clean)

# Glimpse the cleaned dataset
glimpse(hotels_clean)
```

## Create variables for analysis
```{r datawrangling, warning=FALSE, message=FALSE}
hotels_clean <- 
  hotels_clean %>%
# Mutate the year variable to a factor variable
  mutate(
  arrival_date_year = case_when(arrival_date_year==2015 ~ 0,
                            arrival_date_year==2016 ~ 1,
                            arrival_date_year==2017 ~ 2) 
  ) %>%
# Combine the month variable for a season variable
  mutate(
  arrival_date_month = recode(arrival_date_month,
  January = 1,
  February = 2,
  March = 3,
  April = 4,
  May = 5,
  June = 6,
  July = 7,
  August = 8,
  September = 9,
  October = 10,
  November = 11,
  December = 12
)
) %>%
  mutate(arrival_date_season = case_when(arrival_date_month >=3 & arrival_date_month <=5 ~ 0,
                            arrival_date_month >= 6 & arrival_date_month <= 8 ~ 1,
                            arrival_date_month >= 9 & arrival_date_month <= 11 ~ 2,
                            arrival_date_month == 12 | arrival_date_month <= 2 ~ 3)) %>%
# Create a combined variable for number of children and babies due to small numbers
  mutate(childbaby = children + babies)
```

## Label variables
```{r datalabel, warning=FALSE, message=FALSE}
# Lable the factor variables
hotels_clean$is_canceled <- factor(hotels_clean$is_canceled, levels=0:1, labels=c("No", "Yes"))
hotels_clean$arrival_date_year <- factor(hotels_clean$arrival_date_year, levels=0:2, labels=c("2015", "2016", "2017"))
hotels_clean$arrival_date_season <- factor(hotels_clean$arrival_date_season, levels=0:3, labels=c("Spring", "Summer", "Fall", "Winter"))
hotels_clean$is_repeated_guest <- factor(hotels_clean$is_repeated_guest, levels=0:1, labels=c("No", "Yes"))

# Use label() function from table1 package to label the variables
label(hotels_clean$hotel) <- "Type of hotel"
label(hotels_clean$lead_time) <- "Number of days between booking and arrival"
label(hotels_clean$arrival_date_year) <- "Year of arrival date"
label(hotels_clean$arrival_date_season) <- "Season of arrival date"
label(hotels_clean$stays_in_weekend_nights) <- "Number of weekend nights stayed"
label(hotels_clean$stays_in_week_nights) <- "Number of week nights stayed"
label(hotels_clean$adults) <- "Number of adults"
label(hotels_clean$childbaby) <- "Number of children and babies"
label(hotels_clean$is_repeated_guest) <- "Repeated guest"
label(hotels_clean$booking_changes) <- "Number of changes to the booking"
label(hotels_clean$adr) <- "Average daily rate, $"
```

# Results

## Table 1: Characteristics of hotel bookings overall and by cancellation
```{r table1, warning=FALSE, message=FALSE}
table1(~ hotel + lead_time + arrival_date_year + arrival_date_season + 
         stays_in_weekend_nights + stays_in_week_nights + adults + 
         childbaby + is_repeated_guest + booking_changes + adr
         | is_canceled, data=hotels_clean, c(left="Overall"))
```

## Figure 1: Number of completed and cancelled hotel bookings by season and year
```{r figure1, fig.height=8, fig.width=10, fig.align="center", warning=FALSE, results="asis"}
# Define a custom theme to start with
custom_theme <- theme(
  panel.background = element_blank(),
  axis.ticks = element_blank(),
  axis.title = element_text(size=14, face="bold"),
  axis.line = element_line(size = 0.5),
  plot.title = element_text(color="black", size=14, face="bold", hjust=0.5),
)

# Data for figure 1
hotels_summary1 <- 
  hotels_clean %>% 
  group_by(arrival_date_year, arrival_date_season, is_canceled) %>%
  summarize(
    n=n()
  )

# Generate figure 1 using geom_bar()
figure1 <- ggplot(data=hotels_summary1, aes(x=arrival_date_season, y=n, fill=is_canceled)) +
           # Add stacked bar plot
           geom_bar(stat="identity") +
           # Specify color manual
           scale_fill_manual(name = "Cancelled booking",
                             values=c("#40B0A6", "#E1BE6A"),
                             labels = c("No", "Yes")) +
           facet_wrap(~ arrival_date_year, ncol=1) +
           # Add labels
           labs(x="Season",
                y="Number of hotel bookings",
                title="Number of completed and cancelled hotel bookings by season and year",
                subtitle = "Larger number of bookings and cancellations during summer",
                caption = "Kening Jiang") +
           # Modify the theme
           custom_theme +
                 theme(axis.text.x=element_text(size=12),
                       axis.text.y=element_text(size=14),
                       legend.text = element_text(size = 12),
                       legend.title=element_text(size=14, face="bold"),
                       plot.subtitle=element_text(size=12, hjust=0.5),
                       legend.box.background = element_rect(),
                       legend.box.margin = margin(5,5,5,5),
                       strip.text.x = element_text(size = 14, face="bold"),
                       strip.background = element_rect(fill="#D3E1D0"),
                       plot.caption = element_text(size=14)
                 )
# Display the plot
figure1
```

## Figure 2: Number of completed and cancelled hotel bookings by hotel type
```{r figure2, fig.height=8, fig.width=10, fig.align="center", warning=FALSE, results="asis"}
# Data for figure 2
hotels_summary2 <- 
  hotels_clean %>% 
  group_by(is_canceled, hotel) %>%
  summarize(
    n=n()
  )

# Generate figure 2 using geom_bar()
figure2 <- ggplot(data=hotels_summary2, aes(x=hotel, y=n, fill=is_canceled)) +
           # Add stacked bar plot
           geom_bar(stat="identity") +
           # Specify color manual
           scale_fill_manual(name = "Cancelled booking",
                             values=c("#40B0A6", "#E1BE6A"),
                             labels = c("No", "Yes")) +
           # Add labels
           labs(x="Hotel type",
                y="Number of hotel bookings",
                title="Number of completed and cancelled hotel bookings by hotel type",
                subtitle = "Larger number of bookings and cancellations in city hotels",
                caption = "Kening Jiang") +
           # Modify the theme
           custom_theme +
                 theme(axis.text.x=element_text(size=12),
                       axis.text.y=element_text(size=14),
                       legend.text = element_text(size = 12),
                       legend.title=element_text(size=14, face="bold"),
                       plot.subtitle=element_text(size=12, hjust=0.5),
                       legend.box.background = element_rect(),
                       legend.box.margin = margin(5,5,5,5),
                       plot.caption = element_text(size=14)
                 )

# Display the plot
figure2
```

## Figure 3: Characteristics of hotel bookings by cancellation
```{r figure3_1, fig.height=8, fig.width=10, fig.align="center", warning=FALSE, results="asis"}
# Data for figure 3
# Mean of continuous characteristics by cancellation
hotels_summary3_mean <- 
  hotels_clean %>% 
  group_by(is_canceled) %>%
  summarize(
    lead_time=mean(lead_time),
    stays_in_weekend_nights = mean(stays_in_weekend_nights),
    stays_in_week_nights = mean(stays_in_week_nights),
    adults = mean(adults),
    childbaby = mean(childbaby),
    booking_changes = mean(booking_changes),
    adr = mean(adr)
  ) %>%
pivot_longer(cols = c(lead_time, stays_in_weekend_nights, stays_in_week_nights, adults, childbaby, booking_changes, adr), names_to = "measure", values_to = "mean") 

# Standard deviation of continuous characteristics by cancellation
hotels_summary3_sd <- 
  hotels_clean %>% 
  group_by(is_canceled) %>%
  summarize(
    lead_time=sd(lead_time),
    stays_in_weekend_nights = sd(stays_in_weekend_nights),
    stays_in_week_nights = sd(stays_in_week_nights),
    adults = sd(adults),
    childbaby = sd(childbaby),
    booking_changes = sd(booking_changes),
    adr = sd(adr)
  ) %>%
pivot_longer(cols = c(lead_time, stays_in_weekend_nights, stays_in_week_nights, adults, childbaby, booking_changes, adr), names_to = "measure", values_to = "sd") 

# Join for combined dataset for mean and standard deviation of characteristics
hotels_summary3 <- inner_join(hotels_summary3_mean, hotels_summary3_sd, by=c("is_canceled", "measure"))
hotels_summary3_1 <- hotels_summary3 %>% filter(measure=="lead_time" | measure == "adr")
hotels_summary3_2 <- hotels_summary3 %>% filter(measure!="lead_time" & measure != "adr")

# Modify the labels for facet
measure.labs <- c("Time before arrival", "Average daily rate")
names(measure.labs) <- c("lead_time", "adr")

# Generate figure 3 using geom_bar() and geom_errorbar()
figure3_1 <- ggplot(hotels_summary3_1, aes(x=is_canceled, y=mean)) + 
             # Add the bar plot for mean of characteristics
             geom_bar(position=position_dodge(), stat="identity", 
             fill="#E1BE6A") + 
             # Add the error bar for standard deviation of characteristics
             geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=.2) +
             # Create facet by variable
             facet_wrap(~ measure, ncol=2, labeller = labeller(measure = measure.labs))  +
             # Add labels
             labs(x="Bookings cancelled",
                  y="Mean and standard deviation",
                  title="Characteristics of bookings by cancellation",
                  subtitle = "Longer time and higher average daily rate among cancelled bookings",
                  caption = "Kening Jiang") +
             # Modify the theme
              custom_theme +
                 theme(axis.text.x=element_text(size=12),
                       axis.text.y=element_text(size=14),
                       legend.text = element_text(size = 12),
                       legend.title=element_text(size=14, face="bold"),
                       plot.subtitle=element_text(size=12, hjust=0.5),
                       legend.box.background = element_rect(),
                       legend.box.margin = margin(5,5,5,5),
                       strip.text.x = element_text(size = 14, face="bold"),
                       strip.background = element_rect(fill="#D3E1D0"),
                       plot.caption = element_text(size=14)
                 )

# Display the plot
figure3_1
```

```{r figure3_2, fig.height=8, fig.width=10, fig.align="center", warning=FALSE, results="asis"}
# Modify the labels for facet
measure.labs <- c("Number of weekend nights", "Number of weekdays", "Number of adults", "Number of children/babies", "Changes in bookings")
names(measure.labs) <- c("stays_in_weekend_nights", "stays_in_week_nights", "adults", "childbaby", "booking_changes")

# Generate figure 3 using geom_bar() and geom_errorbar()
figure3_2 <- ggplot(hotels_summary3_2, aes(x=is_canceled, y=mean)) + 
             # Add the bar plot for mean of characteristics
             geom_bar(position=position_dodge(), stat="identity", 
             fill="#E1BE6A") + 
             # Add the error bar for standard deviation of characteristics
             geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=.2) +
             # Create facet by variable
             facet_wrap(~ measure, ncol=2, labeller = labeller(measure = measure.labs))  +
             # Add labels
             labs(x="Bookings cancelled",
                  y="Mean and standard deviation",
                  title="Characteristics of bookings by cancellation (continued)",
                  subtitle = "Number of weekdays and weekend days and adult guests are higher among cancelled bookings",
                  caption = "Kening Jiang") +
             # Modify the theme
             custom_theme +
                 theme(axis.text.x=element_text(size=12),
                       axis.text.y=element_text(size=14),
                       legend.text = element_text(size = 12),
                       legend.title=element_text(size=14, face="bold"),
                       plot.subtitle=element_text(size=12, hjust=0.5),
                       legend.box.background = element_rect(),
                       legend.box.margin = margin(5,5,5,5),
                       strip.text.x = element_text(size = 14, face="bold"),
                       strip.background = element_rect(fill="#D3E1D0"),
                       plot.caption = element_text(size=14)
                 )

# Display the plot
figure3_2
```

## Univariate analysis for the association between each characteristic and odds of cancelled booking
```{r univariate, warning=FALSE, message=FALSE}
# Store the name of outcome and continuous and cateogorical variables
outcomes <- hotels_clean %>%
  select(is_canceled) %>% colnames
covars_cont <- hotels_clean %>%
  select(lead_time, stays_in_weekend_nights, stays_in_week_nights, adults, childbaby, booking_changes, adr) %>% colnames
covars_cat <- hotels_clean %>%
  select(hotel, arrival_date_year, arrival_date_season, is_repeated_guest) %>% colnames

# Univariate logistic regression for continuous variables
paramlist1 <- expand_grid(outcomes, covars_cont)
uni_summary1 <- paramlist1 %>% 
  mutate(mod = map2(outcomes, covars_cont, ~ glm(reformulate(.y, .x), data = hotels_clean, family = binomial)),
         res = map(mod, broom::tidy)) 
uni_summary1$res

# Univariate logistic regression for categorical variables
paramlist2 <- expand_grid(outcomes, covars_cat)
uni_summary2 <- paramlist2 %>% 
  mutate(mod = map2(outcomes, covars_cat, ~ glm(reformulate(.y, .x), data = hotels_clean, family = binomial)),
         res = map(mod, broom::tidy)) 
uni_summary2$res
```

```{r univariateplot, warning=FALSE, message=FALSE}
# Use map and map_Dbl() for coefficients of regression models
coef1 <- cbind(covars_cont, outcomes %>% 
  map(~ map(covars_cont, function(.y) glm(reformulate(.y, .x), data = hotels_clean, family = binomial))) %>% 
  flatten %>% 
  map(summary) |> map(.f = coef)|> map_dbl(2) )
summary1 <- outcomes %>% 
  map(~ map(covars_cont, function(.y) glm(reformulate(.y, .x), data = hotels_clean, family = binomial))) %>% 
  flatten %>% 
  map(summary)

# Standard error of regression models
str1 <- vector("numeric", length(covars_cont))
for (i in 1:7){
  str1[i] <- coefficients(summary1[[i]])[2,2]
}

# Create the summary table for univariate analysis
uni_summary <- as.data.frame(cbind(coef1, str1))
uni_summary <- uni_summary %>%
  rename(
    variable = covars_cont,
    coefficients = V2,
    strerr = str1
    ) 
uni_summary$coefficients <- as.numeric(uni_summary$coefficients)
uni_summary$strerr <- as.numeric(uni_summary$strerr)
```

## Figure 4. Summary of univariate analysis
```{r figure4, fig.height=8, fig.width=10, fig.align="center", warning=FALSE, results="asis"}
# Modify x labels for plot
x_labels <- c('Average daily rate', 'N_adults', 'Booking changes', 'N_children', 'Time before arrival', 'N_weekdays', 'N_weekends')

# Generate figure 4 using geom_point() and geom_errorbar()
Figure4 <-ggplot(uni_summary, aes(x=variable, y=coefficients)) +
           # Add points
           geom_point() +
           # Add error bars
           geom_errorbar(aes(ymin=coefficients - 1.96 * strerr, ymax = coefficients + 1.96 * strerr), width=.1) +
           # Add horizontal line at y=0
           geom_hline(yintercept=0, color = "red") +
           # Modify the x labels
           scale_x_discrete(labels=x_labels) +
           # Add labels
           labs(x="Variable",
                y="Log (odds ratio) and 95% confidence interval",
                title="Summary of univariate analysis",
                subtitle = "All the continuous variables are significantly associated with cancellation",
                caption = "Kening Jiang") +
           # Modify the theme
           custom_theme +
                 theme(axis.text.x=element_text(size=12),
                       axis.text.y=element_text(size=14),
                       legend.text = element_text(size = 12),
                       legend.title=element_text(size=14, face="bold"),
                       plot.subtitle=element_text(size=14, hjust=0.5),
                       legend.box.background = element_rect(),
                       legend.box.margin = margin(5,5,5,5),
                       plot.caption = element_text(size=14)
                 )
# Display the plot
Figure4
```

```{r multivariate, warning=FALSE, message=FALSE}
# Multivairable logistic regression 
# Year of booking and repeated guest are dropped as they are not significant in the univariate analysis
mvr <- summary(glm(is_canceled~hotel + lead_time + as.factor(arrival_date_season) + stays_in_weekend_nights
                             + stays_in_week_nights + adults + childbaby + booking_changes + adr, 
                             data = hotels_clean, family = binomial))
mvr

# Summary dataset for coefficients and standard errors of the regression
mvr_summary <- as.data.frame(cbind(mvr$coefficients[2:12, 1], mvr$coefficients[2:12, 2]))
mvr_summary <- mvr_summary %>%
  rename(
    coefficients = V1,
    strerr = V2
    ) 
mvr_summary$variable <- row.names(mvr_summary)  
```

## Figure 5. Summary of multivariable analysis
```{r figure5, fig.height=8, fig.width=10, fig.align="center", warning=FALSE, results="asis"}
# Modify x labels for plot
x_labels2 <- c('Average daily rate', 'N_adults', "Fall", "Summer","Winter", 'Booking changes', 'N_children', "Hotel type", 'Time before arrival', 'N_weekdays', 'N_weekends')

# Generate figure 5 using geom_point() and geom_errorbar()
Figure5 <-ggplot(mvr_summary, aes(x=variable, y=coefficients)) +
           # Add points
           geom_point() +
           # Add error bars
           geom_errorbar(aes(ymin=coefficients - 1.96 * strerr, ymax = coefficients + 1.96 * strerr), width=.1) +
           # Add horizontal line at y=0
           geom_hline(yintercept=0, color = "red") +
           # Modify the x labels
           scale_x_discrete(labels=x_labels2) +
           # Add labels
           labs(x="Variable",
                y="Log (odds ratio) and 95% confidence interval",
                title="Summary of multivariable analysis",
                subtitle = "Season and number of adults and children are not significant",
                caption = "Kening Jiang") +
           # Modify the theme
           custom_theme +
                 theme(axis.text.x=element_text(size=12, angle=90),
                       axis.text.y=element_text(size=14),
                       legend.text = element_text(size = 12),
                       legend.title=element_text(size=14, face="bold"),
                       plot.subtitle=element_text(size=14, hjust=0.5),
                       legend.box.background = element_rect(),
                       legend.box.margin = margin(5,5,5,5),
                       plot.caption = element_text(size=14)
                 )
# Display the plot
Figure5
```

# Summary
Among 2,097 hotel bookings between 2015 and 2017 in the United States, 501 (24%) bookings are cancelled and 1,596 (76%) bookings are completed. We examined whether hotel type, time between booking and arrival, season and year of the bookings, number of weekdays and weekend days, number of adult guests, number of children/babies, changes to the bookings, repeated guest, and average daily rate are associated with cancelled bookings. In multivariable logistic regression, resort (vs. city) hotel, fall (vs. spring) and changes in the bookings are independently associated with lower odds of cancellations, while longer time between booking and arrival, longer stays on weekdays and weekend days, and higher average daily rating are independently associated with higher odds of cancellations.

# Functions used in the analysis

## Data wrangling
* filter()
* select()
* mutate()
* group_by()
* summarise()
* inner_join()
* pivot_longer()

## Purrr functions
* map()
* map_dbl()
* map2()

## Functions from the table1 package in the vignette
* table1()
* label()

## Figures
* geom_bar()
* geom_point()
* geom_errorbar()